---
title: "BLISA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BLISA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  eval = FALSE
)
```

# Introduction

BLISA identifies spatially enriched ligand–receptor interactions from spatial transcriptomics data using local bivariate Moran’s I statistics.

This vignette demonstrates a typical BLISA workflow using Xenium data.

# Load Package

```{r}
library(blisa)
library(scider)
library(CellChat)
library(patchwork)
```

# Load Example Data

Example Xenium dataset distributed with the package:

```{r}
spe <- readRDS(
  system.file("extdata", "xenium380Cell_spe.rds", package = "blisa")
)

dim(spe)
```

# Visualize Spatial Cell Types

```{r}
cell_type_colors <- c(
  RColorBrewer::brewer.pal(7, "Set1"),
  RColorBrewer::brewer.pal(6, "Set2")
)

names(cell_type_colors) <- unique(spe$cell_type)

scider::plotSpatial(
  spe,
  group.by = "cell_type",
  pt.size = 0.2,
  pt.alpha = 0.9,
  cols = cell_type_colors
)
```

# Filter Ligand–Receptor Pairs

Restrict CellChat ligand–receptor database to genes present in the panel:

```{r}
counts_matrix <- SummarizedExperiment::assay(spe, "counts")

LR_pairs_filtered <- filterLRpairs(
  counts = counts_matrix,
  min_ligand = 1,
  min_receptor = 1,
  LR_df = CellChatDB.human$interaction
)

head(LR_pairs_filtered)
```

# Run BLISA

```{r}
system.time(BLISAres<- runBLISA.spe.isolates.removed(
    spe, 
    LR_df = LR_pairs_filtered))

BLISAres$LR_out
```

# Summarize Ligand–Receptor Interaction Enrichment

```{r}
plotLRIsum(BLISAres$LR_out)
```

# Visualize Spatial Distribution of Top LR Pairs

```{r, fig.width=15, fig.height=12}
plots <- lapply(seq_len(nrow(BLISAres$LR_results)), function(i) {
  plotHotspots(BLISAres, index = i)
})

combined_plot <- patchwork::wrap_plots(plots, ncol = 3)

combined_plot

pdf("xenium_LRI_spatial_plots.pdf", width = 15, height = 12)
print(combined_plot)
dev.off()
```

# Run Cell–Cell Interaction Analysis

When `blisa()` is called with a `group` argument (e.g. `group = "cell_type"`),
CCI scores are computed automatically and stored in `BLISAres$CCI_scores`.
To compute or recompute them after the fact, call `runCCI()` on the
`blisa` object and supply `counts_by_group` from `hexBinCells()`:

```{r}
# CCI already computed inside blisa() — retrieve directly
CCIres <- BLISAres$CCI_scores

# Alternatively, compute on an existing blisa object that lacks CCI_scores:
# binned  <- hexBinCells(coords, counts_matrix, bin_size = 50, group = cell_type_vec)
# BLISAres <- runCCI(BLISAres, counts_by_group = binned$counts_by_group)
# CCIres  <- BLISAres$CCI_scores

head(CCIres)
```

# Visualize CCI Heatmap

```{r fig.height=8, fig.width=8}
CCIheatmap(CCIres)
```

```{r, fig.height=10, fig.width=9}
CCIheatmap(CCIres, include_celltypes = c("Invasive_Tumor"))
```

# Visualize One Ligand–Receptor Pair

```{r fig.height=5, fig.width=6}
p <- CCIheatmapOneLR(
  CCIres,
  "CXCL12_CXCR4"
)

draw(
  p,
  heatmap_legend_side = "right",
  padding = unit(c(5, 5, 10, 10), "mm")
)
```

```{r fig.height=7, fig.width=10}
CCIspatial(
  spe,
  BLISAres,
  index = 2
)
```

# Session Information

```{r}
sessionInfo()
```
