---
title: "BLISA Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BLISA Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Introduction

BLISA identifies spatially enriched ligand–receptor interactions from spatial transcriptomics data using local bivariate Moran’s I statistics.

This vignette demonstrates a typical BLISA workflow using Xenium data.

# Load Package

```{r}
library(blisa)
library(scider)
library(CellChat)
library(patchwork)
```

# Load Example Data

Example Xenium dataset distributed with the package:

```{r}
spe <- readRDS(
  system.file("extdata", "xenium380Cell_spe.rds", package = "blisa")
)

dim(spe)
```

# Visualize Spatial Cell Types

```{r}
cell_type_colors <- c(
  RColorBrewer::brewer.pal(7, "Set1"),
  RColorBrewer::brewer.pal(6, "Set2")
)

names(cell_type_colors) <- unique(spe$cell_type)

scider::plotSpatial(
  spe,
  group.by = "cell_type",
  pt.size = 0.2,
  pt.alpha = 0.9,
  cols = cell_type_colors
)
```

# Filter Ligand–Receptor Pairs

Restrict CellChat ligand–receptor database to genes present in the panel:

```{r}
counts_matrix <- SummarizedExperiment::assay(spe, "counts")

LR_pairs_filtered <- filterLRpairs(
  counts = counts_matrix,
  min_ligand = 1,
  min_receptor = 1,
  LR_df = CellChatDB.human$interaction
)

head(LR_pairs_filtered)
```

# Run BLISA

```{r}
system.time(BLISAres<- runBLISA.spe.isolates.removed(
    spe, 
    LR_df = LR_pairs_filtered))

BLISAres$LR_out
```

# Summarize Ligand–Receptor Interaction Enrichment

```{r}
plotLRIsum(BLISAres$LR_out)
```

# Visualize Spatial Distribution of Top LR Pairs

```{r}
plots <- lapply(seq_len(min(6, nrow(BLISAres$LR_out))), function(i) {
  plotLRI.sf(
    BLISAres$bin_sf,
    BLISAres$LR_out,
    index = i
  )
})

patchwork::wrap_plots(plots, ncol = 3)
```

# Run Cell–Cell Interaction Analysis

```{r}
CCIres <- runCCI(
  spe,
  BLISAres
)

head(CCIres)
```

# Visualize CCI Heatmap

```{r fig.height=8, fig.width=8}
CCIheatmap(CCIres)
```

```{r, fig.height=10, fig.width=9}
CCIheatmap(CCIres, include_celltypes = c("Invasive_Tumor"))
```

# Visualize One Ligand–Receptor Pair

```{r fig.height=5, fig.width=6}
p <- CCIheatmapOneLR(
  CCIres,
  "CXCL12_CXCR4"
)

draw(
  p,
  heatmap_legend_side = "right",
  padding = unit(c(5, 5, 10, 10), "mm")
)
```

```{r fig.height=7, fig.width=10}
CCIspatial(
  spe,
  BLISAres,
  index = 2
)
```

# Session Information

```{r}
sessionInfo()
```
